---
title: "Plotting EA Classes"
author: "Emily O'Grady"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cerulean
vignette: >
  %\VignetteIndexEntry{Plotting EA classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Plotting with marea
## A comprehensive guide to visualizing ecosystem data

```{r}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Introduction

The marea package provides a unified plotting interface for ecosystem data in Atlantic Canada. This vignette demonstrates how to create effective visualizations using the built-in plotting functions and how to customize them for your specific needs.

```{r}
library(marea)
library(ggplot2)
library(tidyverse)
```

# Quick Start: Basic Plotting

All marea datasets work the same way - simply use the plot() function, the S3 plotting methods for the EA_data and EA_st classes will give you a default starting point.

```{r}
# Load and plot North Atlantic Oscillation index
data(nao)
plot(nao)
```



# Time Series Data: The ea_data Class

## Available Plot Styles

The ea_data class supports several plotting styles:

### Default Style

Basic line plot with points:

```{r}
plot(nao, style = "default")
```
### Ribbon Style

Shows uncertainty bounds (requires low and high columns):

```{r}
plot(grey_seals, style = "ribbon")
```

### Plain Style

Simple line without points:

```{r}
plot(nao, style = "plain")
```

### Anomaly Style

Bar plot with color-coded positive/negative values:

```{r}
data(nao)

plot(nao, style = "anomaly")
```

### Biomass Style

Special style for biomass data with bold lines and uncertainty:

```{r}
plot(grey_seals, style = "biomass")
```

# Customizing Time Series Plots

Since all plots return ggplot2 objects, you can easily customize them:

```{r}
# Start with base plot
p <- plot(nao, style = "plain")

# Add custom elements
p + 
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.3) +
  labs(
    title = "North Atlantic Oscillation Index",
    subtitle = "With trend line and zero reference",
    caption = "Data source: NOAA"
  ) +
  theme_minimal()
```

# Working with Multiple Variables

Compare different indices:

```{r}
# Load multiple climate indices
data(nao)
data(oni)

# Create comparison plot
nao_recent <- nao$data[nao$data$year >= 2000, ]
oni_recent <- oni$data[oni$data$year >= 2000 & oni$data$month == 1, ]  # January only

# Combine data
combined_data <- data.frame(
  year = nao_recent$year,
  NAO = nao_recent$value,
  ONI = oni_recent$value[match(nao_recent$year, oni_recent$year)]
)

ea_df <- ea_data(
  data = combined_data,
  value_col = "NAO",
  data_type = "Climate Indices Comparison",
  region = "North Atlantic",
  location_descriptor = "Global",
  units = "Index Value"
)
 
plot(ea_df, style = "default") +
  geom_line(aes(y = ONI), color = "blue") +
  labs(
    title = "Climate Indices: NAO vs ONI",
    y = "Index Value",
    x = "Year"
  ) +
  theme_minimal()

```

# Spatial Data: The ea_st Class

Spatial data in marea is represented by the ea_st class, which allows for visualizing geographic data such as temperature, salinity, and other oceanographic variables.

## Loading and subsetting spatial data

```{r}

data(glorys_bottom_temperature)
glorys_2024 <- glorys_bottom_temperature[glorys_bottom_temperature$data$year == 2024 ]

```


## Available Spatial Plot Styles

### Fill Style

Color-coded spatial maps:

```{r}
plot(glorys_2024, style = "fill")
```

### Bubble Style

### Contour Style

Contour lines showing value gradients:

```{r}
plot(glorys_2024, style = "contour")
```

## Customizing Spatial Maps

Add coastlines and improve styling:

```{r}
p <- plot(glorys_2024, style = "fill")

# Customize the map
p + 
  scale_fill_viridis_c(name = "Temperature\n(°C)") +
  labs(
    title = "Bottom Temperature in the Maritimes Region",
    subtitle = "GLORYS Reanalysis Data"
  ) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm")
  )
```

## Temporal Filtering for Spatial Data

Filter spatial data by time periods:

```{r}
# Filter for specific months if time dimension exists
if ("month" %in% names(glorys_bottom_temperature$data)) {
  winter_data <- glorys_2024
  winter_data$data <- winter_data$data[
    glorys_2024$data$month %in% c(12, 1, 2), 
  ]
  
  plot(winter_data, style = "fill") +
    labs(title = "2024 Winter Bottom Temperature")
}
```

# Specialized Visualizations

## Climate Index Comparisons

Create correlation plots:

```{r}
# Example correlation plot (if sufficient data overlap exists)
if (nrow(nao$data) > 20) {
  recent_nao <- nao[nao$data$year >= 1990 ]
  
  # Add trend analysis
  p <- plot(recent_nao, style = 'plain')+
      labs(x = 'Year', y = 'NAO Index')+
      ggtitle("NAO Index Over Time (1990 - Present)") 
  
  # Add regression line
  lm_fit <- lm(value ~ year, data = recent_nao$data)
  p <- p +
    geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
    theme_minimal()

  # Add text with trend
  slope <- round(coef(lm_fit)[2], 4)
  
  p <- p +
    geom_text(aes(x = 2000, y = max(recent_nao$data$value), 
                  label = paste("Trend:", slope, "per year")), 
              color = "red", hjust = 0)

  print(p)
}
```

## Multi-panel Comparisons

This example shows how to compare multiple datasets side by side using base R graphics. This is useful for visualizing trends across different datasets in a single view.


Compare different datasets side by side:

```{r}
# Create a multi-panel plot
par(mfrow = c(2, 2))

# Panel 1: NAO
plot(nao$data$year, nao$data$value, type = "l", 
     main = "North Atlantic Oscillation", 
     xlab = "Year", ylab = "Index")

# Panel 2: Grey Seals
plot(grey_seals$data$year, grey_seals$data$value, type = "l",
     main = "Grey Seal Abundance",
     xlab = "Year", ylab = "Count")

# Panel 3: ONI (subset)
oni_subset <- oni$data[oni$data$month == 1, ]  # January only
plot(oni_subset$year, oni_subset$value, type = "l",
     main = "Oceanic Niño Index (January)",
     xlab = "Year", ylab = "Temperature Anomaly (°C)")

# Panel 4: Bottom Temperature
if (exists("azmp_bottom_temperature")) {
  data(azmp_bottom_temperature)
  bottom_temp_eb <- azmp_bottom_temperature[azmp_bottom_temperature$data$region == 'Emerald Basin']
  plot(bottom_temp_eb$data$year, bottom_temp_eb$data$value, 
       type = "p",
       main = "Emerald Basin Bottom Temperature",
       xlab = "Year", ylab = "Temperature (°C)")
}

# Reset to single panel
par(mfrow = c(1, 1))
```

# Advanced Customization

## Custom Color Schemes

Apply custom color palettes:

```{r}
# Custom color scheme for temperature data
custom_temp_plot <- function(data_obj) {
  p <- plot(data_obj, style = "fill")
  
  p + 
    scale_fill_gradient2(
      low = "blue", mid = "white", high = "red",
      midpoint = mean(data_obj$data$value, na.rm = TRUE),
      name = "Temperature\n(°C)"
    ) +
    theme_minimal()
}

# Apply to spatial data
if (exists("glorys_2024")) {
  custom_temp_plot(glorys_2024)
}
```

## Adding Annotations

Add reference lines and annotations:

```{r}
p <- plot(nao, style = "default")

p + 
  geom_hline(yintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 1960, y = 1.2, label = "Positive phase", 
           hjust = 0, size = 3) +
  annotate("text", x = 1960, y = -1.2, label = "Negative phase", 
           hjust = 0, size = 3) +
  labs(
    title = "North Atlantic Oscillation with Phase Indicators",
    caption = "Dashed lines show ±1 standard deviation"
  )
```

## Working with Uncertainty

### Confidence Intervals

Display and customize uncertainty bounds:

```{r}
# Plot with custom confidence styling
p <- plot(grey_seals, style = "default")

p + 
  geom_ribbon(aes(ymin = low, ymax = high), alpha = 0.3, fill = "lightblue") +
  geom_line( color = "darkblue") +
  geom_point( color = "darkblue") +
  labs(
    title = "Grey Seal Abundance with 95% Confidence Intervals",
    y = "Abundance (thousands)"
  ) +
  theme_bw()
```

## Export and Saving

### High-Quality Output

Save plots for publications:

```{r}
# Create publication-ready plot
pub_plot <- plot(grey_seals, style = "ribbon") +
  labs(
    title = "Grey Seal Population Trends",
    subtitle = "Sable Island, Nova Scotia",
    x = "Year",
    y = "Estimated Abundance (thousands)",
    caption = "Data source: DFO Science"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

# Save as high-resolution PNG
ggsave("grey_seals_trends.png", pub_plot, 
       width = 8, height = 6, dpi = 300)

# Save as PDF for vector graphics
ggsave("grey_seals_trends.pdf", pub_plot, 
       width = 8, height = 6)
```


# Tips and best practices

1. Check Data structure first

```{r}
# Always inspect your data first
print(nao)
summary(nao)
str(nao$data)
```

2. Use Appropriate Styles

- Time series: Use "default" or "ribbon" for most cases
- Anomalies: Use "anomaly" style for climate indices
- Biomass: Use "biomass" style for population data
- Spatial: Use "fill" for continuous fields, "bubble" for point data

3. Customize gradually

```{r}
# Start simple
p1 <- plot(nao)

# Add one element at a time
p2 <- p1 + theme_minimal()
p3 <- p2 + labs(title = "Custom Title")
p4 <- p3 + geom_smooth(method = "loess", se = FALSE)

# Final result
p4
```

4. Consistent styling

```{r}
# Create a custom theme for all your plots
my_theme <- theme_bw() +
  theme(
    plot.title = element_text(size = 14, face = "bold", colour = 'darkblue'),
    axis.title = element_text(size = 12),
    legend.position = "bottom"
  )

# Apply to any plot
plot(nao) + my_theme
```

# Getting Help

## Available datasets

```{r}
# See all available datasets
marea_metadata()
```

## Plot Method Documentation

```{r}
# Get help on plotting functions
?plot.ea_data
?plot.ea_st

# General package help
?marea
```

# Citation

```{r}
# Cite the package in your work
citation("marea")
```

