---
title: "Extracting Data based on Spatial Range"
author: "Emily O'Grady"
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cerulean
vignette: >
  %\VignetteIndexEntry{Spatial Data Extraction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, warnings = FALSE, message = FALSE}
library(marea)
library(tidyverse)
library(sf)
library(cmocean) # for colour scales
library(Mar.data) 
# available from https://github.com/PopulationEcologyDivision/Mar.data
```


## Introduction

In this vignette, we will demonstrate how to extract data from the `marea` package based on a spatial range. This is particularly useful when you want to focus on a specific geographic area for your analysis.

## Useful tools

- `marea` provides curated spatial datasets

- `sf` standard for handling spatial data, coordinate reference systems and spatial filtering

- `tidyverse` for data manipulation

- `ggplot2` for plotting spatial data


## Example: Extracting Data for a Specific Region

### load in a spatial dataset

First we will load in an example of a spatial dataset from marea. For this example we will use glorys bottom temperature. 

```{r}
data("glorys_bottom_temperature")
```

### load in spatial box of interest

We will use an example polygon from Mar.data, but any custom polygon could be created and used here.

```{r}
data("LFAs_sf")

lfa_33 <- LFAs_sf %>%
  filter(LFA == "33") %>%
  st_transform(crs = st_crs(glorys_bottom_temperature@data)) # use same CRS as the data for spatial operations

```

### check for overlap and subset the data

Now we can see where out area of interest overlaps with the data

```{r}

# first we check for overlap
overlap_check <- st_intersects(glorys_bottom_temperature@data, lfa_33, sparse = FALSE)

# How many survey points fall within our study area?
sum(overlap_check)
```

Now we can subset the data to only include points that fall within our area of interest

```{r}
area_subset <- glorys_bottom_temperature@data %>% 
  st_filter(lfa_33)

```


Let's visualize the extraction to verify it worked

```{r}
ggplot() +
  # 1. Base map context (optional): Plot the background grid faintly
  geom_sf(data = glorys_bottom_temperature@data, color = "grey90", size = 0.8, shape = 15) +
  
  # 2. Extracted Data: Map the 'value' column to color. 
  # Using shape=15 (square) to make the dense grid points look like a solid layer
  geom_sf(data = area_subset, aes(color = value), size = 2.5, shape = 15) +
  
  # 3. Top Layer: Our subsetting box (fill = NA ensures we can see the data inside)
  geom_sf(data = lfa_33, fill = NA, color = "black", linewidth = 1.2) +
  
  # Apply the oceanographic cmocean 'thermal' palette for temperature
  scale_color_cmocean(name = "thermal") +
  
  theme_minimal() +
  labs(
    title = "Spatial Subsetting of GLORYS Bottom Temperature",
    subtitle = "Black box: Study Area boundary | Colors: Extracted Reanalysis Data",
    color = "Bottom\nTemp (°C)",
    x = "Longitude", y = "Latitude"
  ) +
  theme(
    legend.position = "right",
    panel.grid.major = element_line(color = "grey95")
  )
```







